---
title: "Working with data"
author: "Jim Leach"
date: "`r Sys.Date()`"
output: 
  revealjs::revealjs_presentation:
    reveal_plugins: ["notes"]
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
```

## In this section

* Data manipulation with `dplyr`
* Getting data in to and out of R - `readr` and `readxl`
* Challenges

# Data manipulation

## `dplyr`

[R for Data Science Chapter](http://r4ds.had.co.nz/transform.html)

A grammar of data manipultion.

```{r example_dplyr, eval = FALSE}
function_name(data, arguments)
```

A consistent approach to most data manipulation challenges:

* `select()`
* `filter()`
* `mutate()`
* `summarise()`
* `arrange()`
* `group_by()`

<aside class="notes">
* Explain each function:
    * mutate() adds new variables that are functions of existing variables
    * select() picks variables based on their names.
    * filter() picks cases based on their values.
    * summarise() reduces multiple values down to a single summary.
    * arrange() changes the ordering of the rows.
* Note similarity to SQL
* (Note about SQL translation?)
</aside>

## data - `storms`

```{r show_packages, eval = TRUE}
library(dplyr)
storms
```

<aside class="notes">
* Built in data set that comes with `ggplot2`
* Contains the prices and other attributes of almost 54,000 diamonds. 
* Diamond characteristics and prices
* Type `?diamonds` at the console to get more details
</aside>

## `select()`

Picks columns based on their name.

```{r select_eg}
select(storms, name, year, month, day, status)
```

* Lots of helper functions, e.g. `starts_with()`, `ends_with()`, see `?select_helpers`

<aside class="notes">
* Remember `?function`
* To drop columns use `-`
* Can use numeric positions of the columns - start from 1 on the left
* Helper functions (?select_helpers) can be used _inside_ `select()`
* Will _only_ select the fields we specify (`rename()` keeps all)
</aside>

## `filter()`

Selects rows based on (a) condition(s).

```{r filter_eg}
filter(storms, status == "hurricane")
```

* Criteria: `==` , `>`, `>=`, `!=` (not equal to), `between()`

<aside class="notes">
* Multiple _and_ criteria can be separated with `&` or `,`
* Multiple _or_ (careful!) separated with `|`
* Use `!` for not
</aside>

## `mutate()`

Adds new columns that are functions of existing columns:

```{r mutate_eg, eval = FALSE} 
mutate(storms, 
       wind_times_pressure = wind * pressure, 
       wind_over_pressure = wind / pressure)
```

```{r mutate_eg_show, echo = FALSE} 
mutate(
  storms, 
  wind_times_pressure = wind * pressure,
  wind_over_pressure = wind / pressure
  ) %>% 
  select(name, year, month, day, wind, pressure, wind_times_pressure, wind_over_pressure)
```

* Lots of useful functions can be used: `case_when()`, `lead()`/`lag()`, `coalesce()`

<aside class="notes">
* The other columns are hidden in the output above
* Add multiple columns using `,`, even using columns you have just created!
* `mutate()` will _keep_ all the other variables, too
* `transmute()` will _drop_ the other variables
* Most operations can be achieved with a `mutate()`
</aside>

## `summarise()`/`summarize()`

Reduces multiple values to a single summary

```{r summarise_eg}
summarise( # Or summarize
  storms, 
  avg_pressure = mean(pressure),
  max_pressure = max(pressure)
  )
```

* Use _any_ summary function: `mean()`, `min()`, `max()`, `n()` (count)
* Multiple values separated with `,`

<aside class="notes">
* Whole table summaries not that useful - we probably want summaries in _groups_
</aside>

## Piping

> ` %>% `

[R for Data Science Chapter](http://r4ds.had.co.nz/pipes.html)

* Read as "then"
* `x %>% f`   is equivalent to `f(x)`
* `x %>% f(y)`   is equivalent to `f(x, y)`
* `x %>% f %>% g %>% h`   is equivalent to `h(g(f(x)))`

<aside class="notes">
* Comes with the `tidyverse` (from `magrittr`)
* Makes it easy to compose a sequence of operations
* The outcome of the left hand side is fed to the first argument of the function on the right
* Works with `dplyr` because the first argument is always data (and most/all packages in the tidyverse)
* Hence why it's easy to learn
</aside>

## `group_by()`

Group data by one or more columns. Often used before `summarise()` or `mutate()`

```{r group_by_eg}
storms %>% 
  group_by(status, year) %>% 
  summarise(max_wind = max(wind))
```

<aside class="notes">
* Specify multiple groups using `,`
* Use `ungroup()` to remove groups
</aside>

## Without `%>%`

```{r group_by_not_piped_eg}
summarise(group_by(storms, status, year), max_wind = max(wind))
```

<aside class="notes">
* Harder to read, especially with longer pipelines
* Functions don't read in a logical order
* Arguments are separated from functions
</aside>

## `arrange()`

Changes the ordering of rows.

```{r arrange_eg}
arrange(storms, name)
```

* Reverse order with `desc()`, specify multiple values with `,`
* Use the `.by_group` argument to arrange _within_ a group

# Getting data in/out of R

## `readr`/`readxl`

[R for Data Science chapter](http://r4ds.had.co.nz/data-import.html)

* `readr` - read text files (`.txt, .csv`, etc)
* `readxl` - read Excel files
* Both use `read_` family of functions, e.g `read_csv()`, `read_xlsx()`, ...

## Reading from CSV

Use the `read_csv()` function. E.g. on example data:

```{r read_mtcars_eg}
library(readr)
my_mtcars <- read_csv(readr_example("mtcars.csv"))
```


<aside class="notes">
* Does everyone know what a CSV/flat file is?
* `readr_example()` is a handy function inside `readr` that provides some
example CSV files to test reading in a CSV
* Column specification tells us what type of data R guessed each column was - we can specify
</aside>

## The results

```{r show_mtcars_eg}
my_mtcars
```

## Specifying column types

We can specify the column types, see `?cols`

```{r read_mtcars_with_spec_eg}
my_mtcars2 <- read_csv(
  readr_example("mtcars.csv"),
  col_types = cols(
    mpg = col_double(),
    cyl = col_integer(),
    disp = col_double(),
    hp = col_integer(),
    drat = col_double(),
    wt = col_double(),
    qsec = col_double(),
    vs = col_integer(),
    am = col_factor(levels = c(0, 1)),
    gear = col_integer(),
    carb = col_integer()
  )
)
```

<aside class="notes">
* E.g. tell it to read am as a factor (categorical data)
* Lots of handy functions for dealing with column type specificaion - `?spec`
</aside>

## The results

```{r show_mtcars_with_spec_eg}
my_mtcars2
```

## Reading from Excel

Use `read_excel()`

```{r show_xl_eg}
library(readxl)
readxl_example("clippy.xls")
```

```{r show_read_xl_eg}
read_excel(readxl_example("clippy.xls"))
```

<aside class="notes">
* Again, we can use a handy example that comes with the package - here we see how that example works
* There is `read_xls()` and `read_xlsx()` if you need to read files without an extension
* We can specify the sheet to read from with `sheet`, and a cell range with `range` arguments
</aside>

## Writing to text file

Use the helpful `write_` functions in `readr`

```{r out_text, eval = FALSE}
write_csv(my_data, "./path/to/file.csv")
write_tsv(my_data, "./path/to/file.tsv")
write_delim(my_data, "./path/to/file.txt", delim = "\")
```

<aside class="notes">
* Sensible to write to text files
    * Pretty portable
    * Easy to read
    * Well-known
    * Non-proprietary
    * Lightweight
</aside>

## Writing to Excel

Don't. 

Ok, go on then:

```{r out_excel, eval = FALSE}
library(writexl)
write_xlsx(my_data, "./path/to/file.xlsx")
```

(You will need to install `writexl` separately using `install.packages("writexl")`).

<aside class="notes">
* Not sensible to write to Excel
    * Excel will mess with your data (e.g. dates)
    * Much harder to control what happens to data in them (because it's so interactive)
    * Not as portable
    * Only easy to read with (specific version of) Excel
    * Proprietary
</aside>

# Challenges

## Data manipulation

* Using `starwars` data and `dplyr`, can you do/answer the following:
    * Can you select all the columns which contain "color" (without listing them)
    * Which character is the heaviest?
    * How many Droids are there?
    * Which planet has the most characters from it?
    * How many of each species is there, and what is the average height of each?
    * Can you calculate BMI ($BMI = \frac{mass}{(height / 100)^2}$), what is Darth Vader's?

<aside class = "notes">
* We'll take some time to do this - 15 or 20 mins
* Everything should be do-able with the functions we've covered just now
* Ask for help if you get stuck, but consult the documentation first - or check the `tidyverse` site...
</aside>

## Plots

* Using `starwars` and `ggplot2` can you plot:
    * the relationship between height and mass? What about for each gender?
    * a histogram or density plot of character heights? Character mass?
    * a count of each species, or homeworld?
    * a summary of character height for each eye colour?

<aside class = "notes">
* We'll take some time to do this - 15 or 20 mins
* First challenge should be do-able following the example code we used for `diamonds`
* Everything else will require a bit of thinking/research - check the ggplot2 cheatsheet (Help > Cheatsheets > Data Vis...)
* Ask me if you're completely stuck, or you need something explained
</aside>

## Getting help

* [`tidyverse`](tidyverse.org) website
* [R for Data Science](http://r4ds.had.co.nz/)
* Rstudio [community](https://community.rstudio.com/)
* `#rstats` on [twitter](https://twitter.com/search?q=%23rstats)
* [Stack Overflow](https://stackoverflow.com/questions/tagged/r) 
* [Jim](mailto:jim.leach@kpmg.co.uk)